---
title: "Supplement: Grass Valley Model"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
    code-links:
      - text: kbvernon/grass-valley
        icon: github
        href: https://github.com/kbvernon/grass-valley
execute: 
  echo: true
  warning: false
  error: false
---

```{r}
#| label: setup
#| include: false
options(dplyr.summarise.inform = FALSE)
```

## Overview

**Goal**: model the distribution of PaleoIndian sites in Grass Valley, NV, under
the constraint that we keep the methods as simple as possible.

**Data**: (i) site data from Nevada SHPO along with additional survey and (ii)
environmental data, mostly cost-distance estimates derived from a DEM and
Campbell's hiking function.

**Method**: Because PaleoIndian sites are extremely rare, we investigate two
methods for handling small samples in species distribution models. The first is
a comparative approach, similar to a hierarchical or joint SDM that does partial
pooling, sharing information across "species." In this case, we use an
interaction term for time period (`Archaic` and `PaleoIndian`) that approximates
a hierarchical model.

The second method is geostatistical, an attempt to squeeze additional
information out of the spatial co-occurrence of the time periods. For sites
$s \in \mathbb{R}^{2}$, we fit a species distribution model using Fixed Rank
Kriging, formally

$$ 
Y(s) = \mu(X(s)) + \omega(s) + \xi(s)
$$

with $\mu(X(s))$ being the mean trend or deterministic response to spatially-
referenced covariates $X$, $\omega(s)$ being the spatial process, and $\xi(s)$
being a spatial noise term. Following Valavi *et al* (2021), we estimate the
mean trend using down-sampled random forest, and the spatial process is
approximated using a spatial basis function with $K$ spatial knots

$$
\omega(s) \equiv \sum_{k=1}^{K} \phi_{k}(s)\eta_{k}
$$

with $\phi_{k}(s)$ being the spatial basis function and $\eta_{k}$ a spatial
weight derived from the data, which we estimate using the same random forest.

## R Preamble

```{r}
#| label: libraries
library(FRK)
library(patchwork)
library(ranger)
library(sf)
library(sfdep)
library(terra)
library(tidyterra)
library(tidyverse)
library(viridis)
```

Plotting defaults:

```{r}
#| label: plot-defaults
#| code-fold: true
theme_set(theme_bw())

graph_gray <- "gray60"

theme_update(
  axis.text = element_text(color = graph_gray, size = rel(0.7)),
  axis.ticks = element_line(color = graph_gray),
  axis.title = element_text(color = graph_gray, size = rel(0.9)),
  legend.background = element_blank(),
  legend.box = element_blank(),
  legend.spacing.x = grid::unit(0.005, "npc"),
  legend.text = element_text(color = graph_gray, size = rel(0.9)),
  panel.grid = element_blank(),
  plot.subtitle = element_text(size = rel(1), margin = margin(b = 2)),
  plot.title = element_text(size = rel(1), margin = margin(b = 4)),
  strip.background = element_blank(),
  strip.text = element_text(hjust = 0, size = rel(0.9), margin = margin(b = 4))
)

gvcols <- function(...) {
  colors <- c(
    "drab" = "#262A10",
    "pink" = "#B287A3",
    "purple" = "#5D2A42",
    "orange" = "#FF9F1C",
    "yellow" = "#FFD046",
    "dark-drab" = "#161A00",
    "dark-pink" = "#5E3E54",
    "dark-purple" = "#400B27",
    "dark-orange" = "#7D4B02",
    "dark-yellow" = "#7C6200",
    "light-drab" = "#8A8D7C",
    "light-pink" = "#E7BCD7",
    "light-purple" = "#B98A9D",
    "light-orange" = "#FFD0B1",
    "light-yellow" = "#FFE7BB"
  )

  x <- unlist(list(...))
  unname(colors[x])
}

# mapping colors to legend labels is unnecessarily complicated...
GuideColorLegend <- ggproto(
  "GuideColorLegend",
  GuideLegend,
  build_labels = function(key, elements, params) {
    n_labels <- length(key$.label)
    if (n_labels < 1) {
      out <- rep(list(zeroGrob()), nrow(key))
      return(out)
    }

    colour <- key$colour %||% key$fill

    # Extract alpha from the key (which includes override.aes modifications)
    alpha <- key$alpha %||% rep(1, length(colour))

    # Apply alpha to the colour
    colour <- alpha(colour, alpha)

    lapply(
      seq_along(key$.label),
      function(i) {
        element_grob(
          elements$text,
          label = key$.label[i],
          colour = colour[i],
          margin_x = TRUE,
          margin_y = TRUE
        )
      }
    )
  }
)

guide_colorlegend <- function(...) {
  new_guide(
    ...,
    name = "ColorLegend",
    super = GuideColorLegend
  )
}
```

Some constants we use throughout.

```{r}
#| label: constants
# rng seed
# United Starship Enterprise NCC-1701 🖖
SEED <- 1701

# random forest
N_TREES <- 1000
N_THREADS <- 24

# buffer for aoi and map extents
BUFFER <- 3500 # meters

# partial dependence constants
N_GRID <- 100
```

There's a lot of sampling in this analysis, so we set an RNG seed:

```{r}
#| label: seed
set.seed(SEED)
```

## Data

Dependent variable:

-   `presence` in {0, 1}, 1 if an archaeological site has been recorded at a
    location, 0 otherwise. This is presence-only data, so the 0s are a
    **quadrature scheme**, background points for sampling environmental
    conditions in the study area.

Independent variables:

-   `streams` cost-distance, or travel time to nearest stream
-   `lake` cost-distance, or travel time to the estimated shoreline of a
    Pleistocene lake
-   `roads` cost-distance, or travel time to nearest road, which is included to
    capture potential sampling bias

Each cost-distance variable is measured in hours.

**Note:** A limitation of the current approach is that we have to make $K$
copies of the background data for the $K$ species being modeled. In this case,
$K=2$, hence repeating `UNION ALL` for Archaic and PaleoIndian quadrature
points.

```{r}
#| label: data
gpkg <- "data/grass-valley.gpkg"

aoi <- read_sf(gpkg, "aoi")

sites <- read_sf(
  gpkg,
  query = paste(
    "SELECT * FROM (",
    "SELECT id, period, geom FROM sites",
    "WHERE ST_Intersects(geom, (SELECT geom FROM aoi))",
    "AND period IN ('Archaic', 'PaleoIndian')",
    "UNION ALL",
    "SELECT id, 'Archaic' AS period, geom FROM quadrature",
    "UNION ALL",
    "SELECT id, 'PaleoIndian' AS period, geom FROM quadrature",
    ") AS s",
    "LEFT JOIN ecology AS e ON s.id = e.id",
    "ORDER BY pa DESC, period DESC"
  )
)

dem <- rast("data/dem-100m.tif")

costs <- rast(list.files("data", pattern = "^cd-", full.names = TRUE))
names(costs) <- c("lake", "roads", "streams")
```

Here is the distribution of sites in Grass Valley.

```{r}
#| label: figure-overview-map
#| code-fold: true
lake <- read_sf(gpkg, "lake")
roads <- read_sf(gpkg, "roads")

bb8 <- aoi |>
  st_buffer(BUFFER) |>
  st_bbox()

slope <- terrain(dem, "slope", unit = "radians")
aspect <- terrain(dem, "aspect", unit = "radians")

hill <- shade(slope, aspect, 30, 45)
hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
hill <- round(hill)
coltab(hill) <- data.frame(value = 1:1000, col = hcl.colors(1000, "Grays"))
hill <- colorize(hill, "rgb")

jittered_sites <- sites |>
  filter(period %in% c("Archaic", "PaleoIndian"), pa == 1) |>
  select(id, period) |>
  st_jitter(amount = 500)

paleo_sites <- jittered_sites |>
  filter(period == "PaleoIndian") |>
  mutate(y = st_coordinates(geom)[, 2, drop = TRUE]) |>
  arrange(desc(y))

paleo_segments <- vector(mode = "list", length = nrow(paleo_sites))

border <- aoi |>
  st_buffer(3000) |>
  st_transform(4326) |>
  st_cast("LINESTRING")

for (i in seq_along(paleo_segments)) {
  g <- paleo_sites[i, ] |>
    st_transform(4326) |>
    st_nearest_points(border)

  # chop off part of the border around this segment endpoint
  # this will ensure non-overlapping labels
  z <- g |>
    st_cast("POINT") |>
    st_sf(geom = _) |>
    st_filter(border, .predicate = st_is_within_distance, dist = 1) |>
    st_buffer(units::set_units(4500, "m"))

  border <- st_difference(border, z)

  paleo_segments[[i]] <- g |> st_transform(26911)

  remove(i, g, z)
}

paleo_segments <- do.call(c, paleo_segments)

xy <- paleo_segments |>
  st_cast("POINT") |>
  st_sf(geom = _) |>
  st_join(aoi |> mutate(g = 1)) |>
  filter(is.na(g)) |>
  st_coordinates() |>
  as_tibble() |>
  rename_with(tolower)

# if the nearest point on the border is to the left of the site,
# then we want the label to be right-aligned, otherwise left-aligned
paleo_labels <- paleo_sites |>
  st_drop_geometry() |>
  select(id) |>
  bind_cols(xy, xo = c(st_coordinates(paleo_sites)[, "X"])) |>
  relocate(x, y, id) |>
  mutate(
    hjust = ifelse(xo - x >= 0, 1, 0),
    nudge_x = ifelse(hjust == 1, -500, 500)
  )

mx <- 5000

plt <- ggplot() +
  geom_spatraster_rgb(data = hill) +
  geom_spatraster(data = dem, alpha = 0.5) +
  scale_fill_hypso_c("utah_1") +
  geom_sf(
    data = lake,
    fill = alpha("lightblue", 0.3),
    color = "gray80",
    linewidth = 0.2
  ) +
  geom_sf(data = roads, color = "gray55") +
  geom_sf(
    data = jittered_sites |> filter(period == "Archaic"),
    color = gvcols("dark-pink"),
    fill = alpha(gvcols("pink"), 0.75),
    shape = 21,
    size = 2.5
  ) +
  geom_sf(data = paleo_segments, linewidth = 0.3, color = "black") +
  geom_sf(
    data = jittered_sites |> filter(period == "PaleoIndian"),
    color = "black",
    fill = "white",
    shape = 24,
    size = 5.7,
    stroke = 0.2
  ) +
  geom_sf(
    data = jittered_sites |> filter(period == "PaleoIndian"),
    color = gvcols("dark-orange"),
    fill = alpha(gvcols("orange"), 0.75),
    shape = 24,
    size = 3
  ) +
  geom_text(
    data = paleo_labels,
    aes(x, y, label = id),
    size = 10 / .pt,
    hjust = paleo_labels[["hjust"]],
    nudge_x = paleo_labels[["nudge_x"]]
  ) +
  annotate(
    "segment",
    x = 505000,
    y = bb8[["ymin"]],
    xend = 545000,
    yend = bb8[["ymin"]],
    color = "gray65"
  ) +
  annotate(
    "segment",
    x = bb8[["xmin"]] - mx,
    y = 4382000,
    xend = bb8[["xmin"]] - mx,
    yend = 4442000,
    color = "gray65"
  ) +
  scale_x_continuous(breaks = c(505000, 545000)) +
  scale_y_continuous(breaks = c(4382000, 4442000)) +
  coord_sf(
    crs = 26911,
    datum = 26911,
    xlim = bb8[c("xmin", "xmax")] + c(-mx, BUFFER),
    ylim = bb8[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme(
    axis.text.y = element_text(angle = 90, hjust = 0.5),
    axis.title = element_blank(),
    legend.position = "none",
    panel.background = element_blank(),
    panel.border = element_blank(),
    plot.margin = ggplot2::margin()
  )

ggsave(
  filename = "figures/overview-map.svg",
  plot = plt,
  width = 4,
  height = 5
)

remove(
  plt,
  slope,
  aspect,
  mx,
  jittered_sites,
  list = ls(pattern = "paleo_")
)
```

::: {#fig-overview-map style="width: 50%; margin: 0 auto;"}
![](/figures/overview-map.svg){width="100%"}

Overview map showing the sites of Archaic (pink circles) and PaleoIndian sites
(orange triangles) in Grass Valley, NV. All points are randomly jittered to mask
true locations.
:::

### Covariate distributions

Here is the distribution of covariates across Archaic, PaleoIndian, and
background locations.

```{r}
#| label: figure-covariates
#| code-fold: true
d <- sites |>
  filter(!(period == "PaleoIndian" & pa == 0)) |>
  st_drop_geometry() |>
  mutate(
    period = ifelse(pa == 0, "Background", period),
    period = fct_relevel(factor(period), "Background")
  ) |>
  select(period, streams, lake, roads) |>
  pivot_longer(
    c(streams, lake, roads),
    names_to = "variable",
    values_to = "value"
  )

plt <- ggplot(d) +
  geom_density(
    aes(value, fill = period),
    color = "transparent",
    linewidth = 0
  ) +
  scale_fill_manual(
    name = NULL,
    values = alpha(gvcols("light-drab", "pink", "orange"), 0.5),
    guide = guide_colorlegend(reverse = TRUE, override.aes = list(alpha = 1))
  ) +
  facet_wrap(
    vars(variable),
    ncol = 3,
    scale = "free",
    labeller = labeller(variable = \(.x) {
      paste("To", str_to_title(.x))
    })
  ) +
  labs(x = "Hours") +
  coord_cartesian(expand = FALSE) +
  theme(
    axis.line.x.bottom = element_line(color = graph_gray),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.justification = c("right", "top"),
    legend.key.height = unit(0.2, "cm"),
    legend.key.spacing.y = unit(0.2, "cm"),
    legend.key.width = unit(0.3, "cm"),
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    panel.border = element_blank(),
    panel.spacing = unit(12, "pt")
  )

ggsave(
  filename = "figures/covariates.svg",
  plot = plt,
  width = 6.3,
  height = 2.5
)

remove(d, plt)
```

::: {#fig-covariates style="width: 90%; margin: 0 auto;"}
![](/figures/covariates.svg){width="100%"}

Distribution of covariates in parameter space for Archaic, PaleoIndian, and
background sites.
:::

### Spatial Basis Functions

To account for spatial interaction, we use the Matérn32 radial basis function

$$ 
\phi_{k}(s) = (1 + \sqrt{3}d_{sk}/\kappa) exp(-\sqrt{3}d_{sk}/\kappa)
$$

evaluated at spatial knots $k$, with $\kappa$ being the range of support for
each function.

```{r}
#| label: spatial-basis-fns
basis_function <- auto_basis(
  data = as_Spatial(sites),
  nres = 2,
  type = "Matern32"
)

spatial_bases <- eval_basis(basis_function, as_Spatial(sites))

# give each column a name with id and resolution
R <- basis_function@df[["res"]]

colnames(spatial_bases) <- sprintf(
  "sbf_%03d_%s",
  seq_len(length(R)),
  R
)

# filter to include only non-zero bases
j <- which(colSums(spatial_bases) > 0)
spatial_bases <- as.matrix(spatial_bases[, j])

remove(R, j)
```

This might look complicated, but the procedure for evaluating $\phi$ over $K$ is
surprisingly simple. First, let us note that $X(s)$ is, in effect, our design
matrix with dimensions $n$ x $m$, for $n$ observations and $m$ covariates. The
spatial basis function $\phi_{k}(s)$, when evaluated over the $K$ knots,
produces an $n$ x $K$ matrix. The advantage of this approach is that a full
covariance matrix across all sites would be $n$ x $n$, but with the spatial
basis, we can approximate that covariance structure with a $n$ x $K$ matrix
where $K << n$. Equally important, the resulting matrix can simply be
concatenated with the design matrix $X(s)$ - `cbind()` in R-speak - thus,
effectively treating the evaluated spatial basis as $K$ additional covariates.
This means no additional optimization is required to fit the model. The default
algorithm is sufficient.

```{r, results='hide'}
#| label: spatial-bases
#| code-fold: true
plt <- show_basis(basis_function) +
  geom_sf(
    data = aoi,
    fill = alpha(gvcols("orange"), 0.3),
    color = gvcols("orange"),
    linewidth = 1
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin()
  )

bb8 <- st_bbox(sites)

dx <- bb8[["xmax"]] - bb8[["xmin"]]
dy <- bb8[["ymax"]] - bb8[["ymin"]]

ggsave(
  filename = "figures/spatial-bases.svg",
  plot = plt,
  width = 4,
  height = 4 * (dy / dx)
)

remove(plt, bb8, dx, dy)
```

::: {#fig-spatial-bases style="width: 50%; margin: 0 auto;"}
![](/figures/spatial-bases.svg){width="100%"}

Range of support and knot (or center) of each Matérn32 radial basis function.
:::

### Collect data

Now we collect the explanatory and spatial basis features into a separate table.

```{r}
#| label: training-data
training_data <- sites |>
  st_drop_geometry() |>
  select(id, pa, period, streams, roads, lake) |>
  mutate(
    pa = factor(pa),
    period = factor(period)
  ) |>
  bind_cols(spatial_bases)

remove(spatial_bases)
```

### Colinearity check

And a quick test to make sure they are not correlated.

```{r}
#| label: collinearity
training_data |>
  select(lake, roads, streams) |>
  cor() |>
  round(3)
```

## Analysis

To account for class-imbalance in the presence and background sites, we use a
down-sampled random forest as recommended Valavi *et al* (2021). This involves
limiting the training data for each tree in the forest to an equal number of
presence and background sites. The large number of trees, in this case N=1000,
ensures proper sampling of the background environment without biasing the
ensemble toward those background sites.

### Down-sampling utilities

We want to ensure that there are an equal number of PaleoIndian, Arhcaic, and
background sites in every tree of the random forest.

```{r}
#| label: down-sampling-funs
case_weights <- function(x) {
  classes <- paste(
    x[["period"]],
    x[["pa"]],
    sep = "-"
  )
  tbl <- table(classes)
  w <- tbl["PaleoIndian-1"] / tbl[classes]
  w <- unname(w)
  i <- which(classes == "PaleoIndian-1")
  w[i] <- 1
  w
}

sample_fraction <- function(x) {
  classes <- paste(
    x[["period"]],
    x[["pa"]],
    sep = "-"
  )
  k <- length(unique(classes))
  (k * sum(classes == "PaleoIndian-1")) / nrow(x)
}
```

This is roughly equivalent to

```{r}
#| label: down-sample-example
set.seed(SEED)

local({
  classes <- paste(
    training_data[["period"]],
    training_data[["pa"]],
    sep = "-"
  )

  class_sample <- sample(
    classes,
    size = sample_fraction(training_data) * nrow(training_data),
    replace = TRUE,
    prob = case_weights(training_data)
  )

  table(class_sample)
})
```

### Residual spatial autocorrelation

We want to test for spatial autocorrelation in the residuals, so we first fit
the full model and then check its residuals.

```{r}
#| label: random-forest
rf_base <- ranger(
  pa ~ lake + roads + streams + period,
  data = training_data,
  num.trees = N_TREES,
  probability = TRUE,
  case.weights = case_weights(training_data),
  sample.fraction = sample_fraction(training_data),
  num.threads = N_THREADS,
  seed = SEED
)
```

Check residuals

```{r}
#| label: morans-i-test
moran_test <- function(.geometry, .residuals) {
  suppressWarnings({
    neighbors <- st_dist_band(.geometry, lower = 0)

    spatial_weights <- st_kernel_weights(
      nb = include_self(neighbors),
      geometry = .geometry,
      kernel = "gaussian"
    )

    global_moran_perm(
      .residuals,
      nb = neighbors,
      wt = spatial_weights
    )
  })
}

observed <- with(training_data, as.numeric(levels(pa))[pa])

estimate <- predict(
  rf_base,
  data = training_data,
  type = "response",
  num.threads = N_THREADS,
  verbose = FALSE
)[["predictions"]][, "1"]

moran_test(
  st_geometry(sites),
  .residuals = observed - estimate
)
```

### Leave-one-out Cross Validation (LOOCV)

The AUC is a common metric used for evaluating the predictive power of SDMs. For
extremely small samples, however, the AUC is ill-advised, especially for LOO-CV
test sets, so we calculate instead the average difference between the
probabilities assigned to the PaleoIndian sites and the background points in the
test set. For a single PaleoIndian site compared to $N$ background points, the
formula is

$$\Delta P = \frac{1}{N}\sum^{N}_{i=1} P(Y=1) - P(Y_{i}=0) $$

The size of $P(Y=1)$ in this case is analogous to the sensitivity of the model,
it's ability to say where sites are located ("true positives" to use the
technical jargon), and $P(Y=1) - P(Y=0)$ is analogous to the specificity of the
model, it's ability to say where sites are *not* located (or "true negatives").
Estimating this for each PaleoIndian sites requires looping through the data,
training a model on all PaleoIndian sites except a focal site (the leave-one-out
site), then predicting the probability of occurrence on the focal site ($P(Y =
1)$) and all background sites ($P(Y_{i}=0)$). To simplify the for-loop, we
define a helper function for fitting a model to training data and evaluating it
against test data.

::: callout-note
Because we are modeling presence-only data, the expressions `P(Y=1)` and 
`P(Y=0)` are more accurately interpreted as *relative likelihoods* rather than
strict probabilities.
:::

```{r}
#| label: evaluate-fn
fit <- function(.formula, .train) {
  ranger(
    as.formula(.formula),
    data = .train,
    num.trees = N_TREES,
    probability = TRUE,
    case.weights = case_weights(.train),
    sample.fraction = sample_fraction(.train),
    num.threads = N_THREADS,
    seed = SEED
  )
}

estimate <- function(.model, .data) {
  yhat <- predict(
    .model,
    data = .data,
    type = "response",
    num.threads = N_THREADS,
    verbose = FALSE
  )

  yhat[["predictions"]][, "1"]
}

evaluate <- function(.formula, .train, .test) {
  rf <- fit(.formula, .train)
  y_hat <- estimate(rf, .test)

  data.frame(
    presence = y_hat[1],
    absence = I(list(y_hat[-1]))
  )
}
```

Here we evaluate the LOO-CV metric for four models:

1.  Base: cost-distance
2.  Spatial: Base + spatial basis functions
3.  Comparative: Base + period
4.  Full: Base + spatial basis functions + period

Now we specify our models

```{r}
#| label: formulas
sbf_variables <- paste(
  grep("^sbf_", names(training_data), value = TRUE),
  collapse = " + "
)

formulas <- list(
  Base = "pa ~ streams + lake + roads",
  Spatial = paste0("pa ~ streams + lake + roads + ", sbf_variables),
  Comparative = "pa ~ streams + lake + roads + period",
  Full = paste0("pa ~ streams + lake + roads + period + ", sbf_variables)
)

remove(sbf_variables)
```

And evaluate them

```{r}
#| label: loocv
ids <- with(training_data, id[pa == 1 & period == "PaleoIndian"])

model_evals <- vector(mode = "list", length = length(ids))
names(model_evals) <- ids

for (id in ids) {
  test <- bind_rows(
    training_data |> filter(id == {{ id }}),
    training_data |> filter(pa == 0, period == "PaleoIndian")
  )

  train <- training_data |> filter(id != {{ id }})

  i <- with(train, which(period == "PaleoIndian"))

  all_tests <- rbind(
    evaluate(formulas[["Base"]], train[i, ], test),
    evaluate(formulas[["Spatial"]], train[i, ], test),
    evaluate(formulas[["Comparative"]], train, test),
    evaluate(formulas[["Full"]], train, test)
  )

  all_tests[["model"]] <- names(formulas)
  all_tests[["id"]] <- id

  model_evals[[id]] <- all_tests[c("id", "model", "presence", "absence")]
}

model_evals <- do.call(rbind, model_evals)
rownames(model_evals) <- NULL

remove(id, ids, train, test, all_tests)
```

## Results

To help interpret the results of the analysis, we provide all of the following:

1.  LOOCV distributions across sites and models
2.  Partial Dependence Profiles (PDP)
3.  Spatial predictions

These are evaluated for each of the four models detailed above.

### Leave-one-out Cross Validation

We evaluate the ability of each model to discriminate presence from background
locations as described above.

```{r}
#| label: figure-loocv
#| code-fold: true
dp <- model_evals |>
  unnest(absence) |>
  mutate(
    dp = presence - absence,
    id = gsub("26LA", "", id),
    model = fct_relevel(factor(model), "Full", "Comparative", "Spatial")
  ) |>
  group_by(id, model) |>
  summarize(dp = mean(dp), n = length(absence)) |>
  ungroup() |>
  mutate(id = fct_reorder(factor(id), dp, max))

segments <- dp |>
  group_by(id) |>
  summarize(x = min(dp), xend = max(dp)) |>
  ungroup()

plt1 <- ggplot() +
  geom_segment(
    data = segments,
    aes(x, id, xend = xend, yend = id),
    color = graph_gray,
    linewidth = 0.3
  ) +
  geom_point(
    data = dp,
    aes(dp, id, color = model),
    size = 2
  ) +
  scale_color_manual(
    values = gvcols("light-drab", "pink", "orange", "yellow"),
    guide = "none"
  ) +
  labs(x = "P(Y=1) - P(Y=0)", title = "A. Site Distributions") +
  theme(axis.title.y = element_blank())

labels <- dp |>
  group_by(model) |>
  summarize(x = quantile(dp, probs = 0.75)) |>
  ungroup()

plt2 <- ggplot() +
  geom_boxplot(
    data = dp,
    aes(dp, model, fill = model),
    linewidth = 0.3
  ) +
  geom_point(
    data = dp,
    aes(dp, model),
    size = 2,
    shape = 21,
    color = "black",
    fill = "white",
    stroke = 0.2
  ) +
  geom_text(
    data = labels,
    aes(x, model, label = model, color = model),
    nudge_x = 0.01,
    nudge_y = 0.2,
    hjust = 0,
    vjust = 0,
    size = 12 / .pt
  ) +
  scale_fill_manual(
    values = gvcols("light-drab", "pink", "orange", "yellow")
  ) +
  scale_color_manual(
    values = gvcols("drab", "dark-pink", "dark-orange", "dark-yellow")
  ) +
  labs(x = "P(Y=1) - P(Y=0)", title = "B. Model Distributions") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none"
  )

plt <- plt1 + plt2 + plot_layout(axis_titles = "collect")

ggsave(
  filename = "figures/loocv-results.svg",
  plot = plt,
  width = 7,
  height = 4
)

remove(dp, segments, labels, plt, plt1, plt2)
```

::: {#fig-loocv-results style="width: 90%; margin: 0 auto;"}
![](/figures/loocv-results.svg){width="100%"}

Results of leave-one-out cross validation. (A) Distribution of differences
across models for each PaleoIndian site. (B) Distribution of differences across
PaleoIndian sites for each model.
:::

**Spatial sorting bias.** Note that the above LOO-CV currently ignores potential
spatial sorting bias in the estimate, including the difference no matter the
distance between the target PaleoIndian site and the background location.
However, Tobler's Law implies that differences in probability at presence and
background locations should increase with distance, all else being equal. Here
we provide a visual check of this possibility, which also helps to interrogate
the behavior of the models.

```{r}
#| label: figure-loocv-distance
#| code-fold: true
D <- st_distance(
  sites |> filter(pa == 1, period == "PaleoIndian"),
  sites |> filter(pa == 0, period == "PaleoIndian")
)

D <- D / 1000
D <- units::drop_units(D)

ids <- with(sites, id[pa == 1 & period == "PaleoIndian"])

rownames(D) <- ids
colnames(D) <- with(sites, id[pa == 0 & period == "PaleoIndian"])

r <- seq(5, 65, by = 10)

for (i in seq_len(nrow(model_evals))) {
  id <- model_evals[["id"]][i]
  dp <- c()

  for (threshold in r) {
    j <- which(c(D[id, ]) <= threshold)
    j <- unname(j)

    p_1 <- model_evals[i, "presence", drop = TRUE]
    p_0 <- model_evals[i, "absence", drop = TRUE][[1]][j]

    dp <- c(dp, mean(p_1 - p_0))
  }

  model_evals[["range"]][i] <- list(dp)
}

M <- as.data.frame(do.call(rbind, model_evals[["range"]]))
colnames(M) <- r

M <- cbind(model_evals[c("id", "model")], M)

M <- M |>
  pivot_longer(c(-id, -model), names_to = "distance", values_to = "value") |>
  mutate(
    distance = as.numeric(distance),
    model = fct_relevel(factor(model), "Full", "Comparative", "Spatial")
  )

plt <- ggplot(M, aes(distance, value, group = model, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 0.8) +
  scale_color_manual(
    name = "Model",
    values = gvcols("light-drab", "pink", "orange", "yellow"),
    guide = guide_colorlegend(reverse = TRUE)
  ) +
  facet_wrap(vars(id)) +
  labs(x = "Distance [km]", y = "P(Y=1) - P(Y=0)") +
  scale_x_continuous(breaks = seq(5, 65, by = 20)) +
  theme(
    legend.position = "top",
    legend.margin = margin(),
    legend.title = element_text(margin = margin(r = 6))
  )

ggsave(
  plot = plt,
  filename = "figures/loocv-distance.svg",
  width = 6.5,
  height = 7
)

remove(i, j, id, ids, dp, r, p_1, p_0, D, M, plt)
```

::: {#fig-loocv-distance style="width: 80%; margin: 0 auto;"}
![](/figures/loocv-distance.svg){width="100%"}

Results of leave-one-out cross validation compared to background points at
different distances from each left out PaleoIndian site.
:::

### Partial Dependence Profile (PDP)

```{r}
#| label: partials
pdp <- function(variable, model, data) {
  # observations of focal variable
  X <- data[[variable]]
  X <- data.frame(x = seq(min(X), max(X), length = N_GRID))
  names(X) <- variable

  # expand grid
  data[[variable]] <- NULL
  new_data <- X |>
    merge(data, all = TRUE) |>
    subset(period == "PaleoIndian")

  # estimate across expanded grid and average across each value of X
  new_data[["yhat"]] <- estimate(model, new_data)
  new_data <- new_data[c(variable, "yhat")]
  names(new_data) <- c("x", "yhat")
  new_data <- aggregate(
    yhat ~ x,
    data = new_data,
    FUN = mean,
    simplify = TRUE
  )

  # return
  new_data[["variable"]] <- variable
  new_data[c("variable", "x", "yhat")]
}

i <- which(training_data[["period"]] == "PaleoIndian")

models <- Map(
  fit,
  .formula = formulas,
  .train = list(
    training_data[i, ],
    training_data[i, ],
    training_data,
    training_data
  ),
  USE.NAMES = TRUE
)

partial_dependence <- Map(
  pdp,
  variable = rep(c("lake", "streams", "roads"), each = length(models)),
  model = models,
  MoreArgs = list(data = training_data),
  USE.NAMES = FALSE
)

partial_dependence <- do.call(rbind, partial_dependence)

partial_dependence[["model"]] <- rep(names(models), each = N_GRID)
```

```{r}
#| label: figure-partials
#| code-fold: true
partial_dependence[["model"]] <- fct_relevel(
  partial_dependence[["model"]],
  "Full",
  "Comparative",
  "Spatial"
)

plt <- ggplot(partial_dependence, aes(x, yhat, color = model)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(
    name = "Model",
    values = gvcols("light-drab", "pink", "orange", "yellow"),
    guide = guide_colorlegend(reverse = TRUE)
  ) +
  facet_wrap(
    vars(variable),
    ncol = 3,
    scales = "free_x",
    labeller = labeller(variable = \(.x) {
      paste("To", str_to_title(.x))
    })
  ) +
  labs(x = "Hours", y = "P(Y=1)") +
  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) +
  theme(
    legend.position = "bottom",
    legend.margin = margin(),
    legend.title = element_text(margin = margin(r = 6))
  )

ggsave(
  plot = plt,
  filename = "figures/partials.svg",
  width = 6.3,
  height = 2.9
)

remove(plt)
```

::: {#fig-partials style="width: 90%; margin: 0 auto;"}
![](/figures/partials.svg){width="100%"}

Partial dependence plots. Response is technically the relative likelihood of
occurrence of a PaleoIndian site, not a strict probability.
:::

### Predictions

For spatial prediction, we need a raster for each covariate. We have that for
the explanatory variables, but the spatial basis functions require that we apply
Matérn32 to the distance from each grid cell to each knot, as we did with the
presence and background locations.

```{r}
#| label: spatial-covariates
# time period raster (the dummy variable)
r_period <- setValues(dem, "PaleoIndian")
names(r_period) <- "period"

# function to build a raster for each spatial basis function
sbf_to_raster <- function(x, r) {
  xy <- xyFromCell(r, cells(r))
  M <- eval_basis(x, xy)

  # build a raster for each knot (or centroid for spatial basis function)
  rr <- vector(mode = "list", length = ncol(M))
  for (i in seq_len(ncol(M))) {
    r[cells(r)] <- M[, i, drop = TRUE]
    rr[[i]] <- r
  }

  # stack the list of rasters into a single raster with multiple layers
  rast(rr)
}

r_spatial <- sbf_to_raster(basis_function, dem)
names(r_spatial) <- grep("^sbf_", names(training_data), value = TRUE)

# stack all covariate rasters and mask just to make sure we don't predict
# outside of the project area
r_covariates <- mask(c(r_period, costs, r_spatial), dem)

remove(r_period, sbf_to_raster, r_spatial)
```

Now we feed these covariate rasters into our prediction function.

```{r}
#| label: predictions
r_predictions <- lapply(
  models,
  \(rf) {
    predict(r_covariates, rf, fun = estimate)
  }
)

r_predictions <- rast(r_predictions)
names(r_predictions) <- names(formulas)
```

```{r}
#| label: figure-predictions
#| code-fold: true
r_predictions <- mask(r_predictions, dem)

plt <- ggplot() +
  geom_spatraster_rgb(data = hill) +
  geom_spatraster(data = r_predictions, alpha = 0.5) +
  scale_fill_distiller(
    palette = "YlOrBr",
    direction = 1,
    na.value = "transparent"
  ) +
  facet_wrap(
    vars(lyr),
    nrow = 1
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    plot.margin = margin()
  )

ggsave(
  filename = "figures/predictions.svg",
  plot = plt,
  width = 7.5,
  height = 3.2
)
```

::: {#fig-predictions style="width: 100%; margin: 0 auto;"}
![](/figures/predictions.svg){width="100%"}

PaleoIndian response in geographic space.
:::

## Session Info

```{r}
#| label: session-info
# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh[["platform"]][["quarto"]] <- paste(
  quarto::quarto_version(),
  "@",
  quarto::quarto_path()
)

# print it out
pkg_sesh
```