---
title: "Supplement: Grass Valley Model"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
    code-links:
      - text: kbvernon/grass-valley
        icon: github
        href: https://github.com/kbvernon/grass-valley
execute: 
  echo: true
  warning: false
  error: false
---

```{r}
#| label: setup
#| include: false
options(dplyr.summarise.inform = FALSE)
```

## Overview

**Goal**: model the distribution of PaleoIndian sites in Grass Valley, NV, under
the constraint that we keep the methods as simple as possible.

**Data**: (i) site data from Nevada SHPO along with additional survey and (ii)
environmental data, mostly cost-distance estimates derived from a DEM and
Campbell's hiking function.

**Method**: Because PaleoIndian sites are extremely rare, we investigate two
methods for handling small samples in species distribution models. The first is
a comparative approach, similar to a hierarchical or joint SDM that does partial
pooling, sharing information across "species." In this case, we use an
interaction term for time period (`Archaic` and `PaleoIndian`) that approximates
a hierarchical model.

The second method is geostatistical, an attempt to squeeze additional
information out of the spatial co-occurrence of the time periods. For locations
$s \in \mathcal{D} \subset \mathbb{R}^{2}$, we model the probability that a site
is present given spatially-referenced covariates $X$ observed at $s$

$$
P(Y(s)=1 \,|\, X(s)) = f(X(s))
$$

estimating $f$ with a down-sampled classification random forest, following
Valavi *et al* (2021). To let the model leverage spatial covariance among sites,
we then condition the outcome on a spatial basis function $\phi(s)$ evaluated
around $K$ spatial knots

$$
P(Y(s)=1 \,|\, X(s), \phi(s)) = f(X(s), \phi(s))
$$

which we fit by appending the values of $\phi(s)$ to the covariate matrix $X(s)$
as additional features. The Gaussian version of this method, which learns a
linear combination of the spatial bases, is known as Fixed Rank Kriging (FRK).
Random forest instead learns an arbitrary function of them, including their
interaction with the other covariates.

## R Preamble

```{r}
#| label: libraries
library(FRK)
library(patchwork)
library(ranger)
library(sf)
library(sfdep)
library(terra)
library(tidyterra)
library(tidyverse)
library(viridis)
```

Plotting defaults:

```{r}
#| label: plot-defaults
#| code-fold: true
theme_set(theme_bw())

graph_gray <- "gray60"

theme_update(
  axis.text = element_text(color = graph_gray, size = rel(0.7)),
  axis.ticks = element_line(color = graph_gray),
  axis.title = element_text(color = graph_gray, size = rel(0.9)),
  legend.background = element_blank(),
  legend.box = element_blank(),
  legend.spacing.x = grid::unit(0.005, "npc"),
  legend.text = element_text(color = graph_gray, size = rel(0.9)),
  panel.grid = element_blank(),
  plot.subtitle = element_text(size = rel(1), margin = margin(b = 2)),
  plot.title = element_text(size = rel(1), margin = margin(b = 4)),
  strip.background = element_blank(),
  strip.text = element_text(hjust = 0, size = rel(0.9), margin = margin(b = 4))
)

gvcols <- function(...) {
  colors <- c(
    "drab" = "#262A10",
    "pink" = "#B287A3",
    "purple" = "#5D2A42",
    "orange" = "#FF9F1C",
    "yellow" = "#FFD046",
    "dark-drab" = "#161A00",
    "dark-pink" = "#5E3E54",
    "dark-purple" = "#400B27",
    "dark-orange" = "#7D4B02",
    "dark-yellow" = "#7C6200",
    "light-drab" = "#8A8D7C",
    "light-pink" = "#E7BCD7",
    "light-purple" = "#B98A9D",
    "light-orange" = "#FFD0B1",
    "light-yellow" = "#FFE7BB"
  )

  x <- unlist(list(...))
  unname(colors[x])
}

# mapping colors to legend labels is unnecessarily complicated...
GuideColorLegend <- ggproto(
  "GuideColorLegend",
  GuideLegend,
  build_labels = function(key, elements, params) {
    n_labels <- length(key$.label)
    if (n_labels < 1) {
      out <- rep(list(zeroGrob()), nrow(key))
      return(out)
    }

    colour <- key$colour %||% key$fill

    # Extract alpha from the key (which includes override.aes modifications)
    alpha <- key$alpha %||% rep(1, length(colour))

    # Apply alpha to the colour
    colour <- alpha(colour, alpha)

    lapply(
      seq_along(key$.label),
      function(i) {
        element_grob(
          elements$text,
          label = key$.label[i],
          colour = colour[i],
          margin_x = TRUE,
          margin_y = TRUE
        )
      }
    )
  }
)

guide_colorlegend <- function(...) {
  new_guide(
    ...,
    name = "ColorLegend",
    super = GuideColorLegend
  )
}
```

Some constants we use throughout.

```{r}
#| label: constants
# rng seed
# United Starship Enterprise NCC-1701 🖖
SEED <- 1701

# random forest
N_TREES <- 1000
N_THREADS <- 24

# buffer for aoi and map extents
BUFFER <- 3500 # meters

# partial dependence constants
N_GRID <- 100
```

There's a lot of sampling in this analysis, so we set an RNG seed:

```{r}
#| label: seed
set.seed(SEED)
```

## Data

Dependent variable:

-   `presence` in {0, 1}, 1 if an archaeological site has been recorded at a
    location, 0 otherwise. This is presence-only data, so the 0s are a
    **quadrature scheme**, background points for sampling environmental
    conditions in the study area.

Independent variables:

-   `streams` cost-distance, or travel time to nearest stream
-   `lake` cost-distance, or travel time to the estimated shoreline of a
    Pleistocene lake
-   `roads` cost-distance, or travel time to nearest road, which is included to
    capture potential sampling bias

Each cost-distance variable is measured in hours.

```{r}
#| label: data
gpkg <- "data/grass-valley.gpkg"

aoi <- read_sf(gpkg, "aoi")

sites <- read_sf(
  gpkg,
  query = paste(
    "SELECT * FROM (",
    "SELECT id, period, geom FROM sites",
    "WHERE ST_Intersects(geom, (SELECT geom FROM aoi))",
    "AND period IN ('Archaic', 'PaleoIndian')",
    "UNION ALL",
    "SELECT id, 'Archaic' AS period, geom FROM quadrature",
    "UNION ALL",
    "SELECT id, 'PaleoIndian' AS period, geom FROM quadrature",
    ") AS s",
    "LEFT JOIN ecology AS e ON s.id = e.id",
    "ORDER BY pa DESC, period DESC"
  )
)

dem <- rast("data/dem-100m.tif")

costs <- rast(list.files("data", pattern = "^cd-", full.names = TRUE))
names(costs) <- c("lake", "roads", "streams")
```

Here is the distribution of sites in Grass Valley.

```{r}
#| label: figure-overview-map
#| code-fold: true
lake <- read_sf(gpkg, "lake")
roads <- read_sf(gpkg, "roads")

bb8 <- aoi |>
  st_buffer(BUFFER) |>
  st_bbox()

slope <- terrain(dem, "slope", unit = "radians")
aspect <- terrain(dem, "aspect", unit = "radians")

hill <- shade(slope, aspect, 30, 45)
hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
hill <- round(hill)
coltab(hill) <- data.frame(value = 1:1000, col = hcl.colors(1000, "Grays"))
hill <- colorize(hill, "rgb")

jittered_sites <- sites |>
  filter(period %in% c("Archaic", "PaleoIndian"), pa == 1) |>
  select(id, period) |>
  st_jitter(amount = 500)

paleo_sites <- jittered_sites |>
  filter(period == "PaleoIndian") |>
  mutate(y = st_coordinates(geom)[, 2, drop = TRUE]) |>
  arrange(desc(y))

paleo_segments <- vector(mode = "list", length = nrow(paleo_sites))

border <- aoi |>
  st_buffer(3000) |>
  st_transform(4326) |>
  st_cast("LINESTRING")

for (i in seq_along(paleo_segments)) {
  g <- paleo_sites[i, ] |>
    st_transform(4326) |>
    st_nearest_points(border)

  # chop off part of the border around this segment endpoint
  # this will ensure non-overlapping labels
  z <- g |>
    st_cast("POINT") |>
    st_sf(geom = _) |>
    st_filter(border, .predicate = st_is_within_distance, dist = 1) |>
    st_buffer(units::set_units(4500, "m"))

  border <- st_difference(border, z)

  paleo_segments[[i]] <- g |> st_transform(26911)

  remove(i, g, z)
}

paleo_segments <- do.call(c, paleo_segments)

xy <- paleo_segments |>
  st_cast("POINT") |>
  st_sf(geom = _) |>
  st_join(aoi |> mutate(g = 1)) |>
  filter(is.na(g)) |>
  st_coordinates() |>
  as_tibble() |>
  rename_with(tolower)

# if the nearest point on the border is to the left of the site,
# then we want the label to be right-aligned, otherwise left-aligned
paleo_labels <- paleo_sites |>
  st_drop_geometry() |>
  select(id) |>
  bind_cols(xy, xo = c(st_coordinates(paleo_sites)[, "X"])) |>
  relocate(x, y, id) |>
  mutate(
    hjust = ifelse(xo - x >= 0, 1, 0),
    nudge_x = ifelse(hjust == 1, -500, 500)
  )

mx <- 5000

plt <- ggplot() +
  geom_spatraster_rgb(data = hill) +
  geom_spatraster(data = dem, alpha = 0.5) +
  scale_fill_hypso_c("utah_1") +
  geom_sf(
    data = lake,
    fill = alpha("lightblue", 0.3),
    color = "gray80",
    linewidth = 0.2
  ) +
  geom_sf(data = roads, color = "gray55") +
  geom_sf(
    data = jittered_sites |> filter(period == "Archaic"),
    color = gvcols("dark-pink"),
    fill = alpha(gvcols("pink"), 0.75),
    shape = 21,
    size = 2.5
  ) +
  geom_sf(data = paleo_segments, linewidth = 0.3, color = "black") +
  geom_sf(
    data = jittered_sites |> filter(period == "PaleoIndian"),
    color = "black",
    fill = "white",
    shape = 24,
    size = 5.7,
    stroke = 0.2
  ) +
  geom_sf(
    data = jittered_sites |> filter(period == "PaleoIndian"),
    color = gvcols("dark-orange"),
    fill = alpha(gvcols("orange"), 0.75),
    shape = 24,
    size = 3
  ) +
  geom_text(
    data = paleo_labels,
    aes(x, y, label = id),
    size = 10 / .pt,
    hjust = paleo_labels[["hjust"]],
    nudge_x = paleo_labels[["nudge_x"]]
  ) +
  annotate(
    "segment",
    x = 505000,
    y = bb8[["ymin"]],
    xend = 545000,
    yend = bb8[["ymin"]],
    color = "gray65"
  ) +
  annotate(
    "segment",
    x = bb8[["xmin"]] - mx,
    y = 4382000,
    xend = bb8[["xmin"]] - mx,
    yend = 4442000,
    color = "gray65"
  ) +
  scale_x_continuous(breaks = c(505000, 545000)) +
  scale_y_continuous(breaks = c(4382000, 4442000)) +
  coord_sf(
    crs = 26911,
    datum = 26911,
    xlim = bb8[c("xmin", "xmax")] + c(-mx, BUFFER),
    ylim = bb8[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme(
    axis.text.y = element_text(angle = 90, hjust = 0.5),
    axis.title = element_blank(),
    legend.position = "none",
    panel.background = element_blank(),
    panel.border = element_blank(),
    plot.margin = ggplot2::margin()
  )

ggsave(
  filename = "figures/overview-map.svg",
  plot = plt,
  width = 4,
  height = 5
)

remove(
  plt,
  slope,
  aspect,
  mx,
  jittered_sites,
  list = ls(pattern = "paleo_")
)
```

::: {#fig-overview-map style="width: 50%; margin: 0 auto;"}
![](/figures/overview-map.svg){width="100%"}

Overview map showing the sites of Archaic (pink circles) and PaleoIndian sites
(orange triangles) in Grass Valley, NV. All points are randomly jittered to mask
true locations.
:::

### Covariate distributions

Here is the distribution of covariates across Archaic, PaleoIndian, and
background locations.

```{r}
#| label: figure-covariates
#| code-fold: true
d <- sites |>
  filter(!(period == "PaleoIndian" & pa == 0)) |>
  st_drop_geometry() |>
  mutate(
    period = ifelse(pa == 0, "Background", period),
    period = fct_relevel(factor(period), "Background")
  ) |>
  select(period, streams, lake, roads) |>
  pivot_longer(
    c(streams, lake, roads),
    names_to = "variable",
    values_to = "value"
  )

plt <- ggplot(d) +
  geom_density(
    aes(value, fill = period),
    color = "transparent",
    linewidth = 0
  ) +
  scale_fill_manual(
    name = NULL,
    values = alpha(gvcols("light-drab", "pink", "orange"), 0.5),
    guide = guide_colorlegend(reverse = TRUE, override.aes = list(alpha = 1))
  ) +
  facet_wrap(
    vars(variable),
    ncol = 3,
    scale = "free",
    labeller = labeller(variable = \(.x) {
      paste("To", str_to_title(.x))
    })
  ) +
  labs(x = "Hours") +
  coord_cartesian(expand = FALSE) +
  theme(
    axis.line.x.bottom = element_line(color = graph_gray),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.justification = c("right", "top"),
    legend.key.height = unit(0.2, "cm"),
    legend.key.spacing.y = unit(0.2, "cm"),
    legend.key.width = unit(0.3, "cm"),
    legend.position = "inside",
    legend.position.inside = c(1, 1),
    panel.border = element_blank(),
    panel.spacing = unit(12, "pt")
  )

ggsave(
  filename = "figures/covariates.svg",
  plot = plt,
  width = 6.3,
  height = 2.5
)

remove(d, plt)
```

::: {#fig-covariates style="width: 90%; margin: 0 auto;"}
![](/figures/covariates.svg){width="100%"}

Distribution of covariates in parameter space for Archaic, PaleoIndian, and
background sites.
:::

### Spatial Basis Functions

To account for spatial interaction, we use the Matérn32 radial basis function

$$
\phi_{k}(s) = \left(1 + \sqrt{3}d_{sk}/\kappa\right)\exp\left(-\sqrt{3}d_{sk}/\kappa\right)
$$

with $d_{sk}$ being the distance from location $s$ to knot $k$ and $\kappa$ the
range of support. We evaluate this function at two spatial resolutions, giving
$K$ knots in total. Matérn32 is smooth and continuous with infinite support,
unlike discontinuous alternatives such as bi-square, which return zeroes beyond
$\kappa$ and, in our experiments, made random forest extremely sensitive to
$\kappa$ while introducing circular artifacts into geographic estimates.

```{r}
#| label: spatial-basis-fns
basis_function <- auto_basis(
  data = as_Spatial(sites),
  nres = 2,
  type = "Matern32"
)

spatial_bases <- eval_basis(basis_function, as_Spatial(sites))

# give each column a name with id and resolution
R <- basis_function@df[["res"]]

colnames(spatial_bases) <- sprintf(
  "sbf_%03d_%s",
  seq_len(length(R)),
  R
)

# filter to include only non-zero bases
j <- which(colSums(spatial_bases) > 0)
spatial_bases <- as.matrix(spatial_bases[, j])

remove(R, j)
```

The procedure for evaluating $\phi$ over $K$ is simpler than it might look.
First, let us note that $X(s)$ is, in effect, our design matrix with dimensions
$n$ x $m$, for $n$ observations and $m$ covariates. The spatial basis function
$\phi_{k}(s)$, when evaluated over the $K$ knots, produces an $n$ x $K$ matrix.
The advantage of this approach is that a full covariance matrix across all sites
would be $n$ x $n$, but with the spatial basis, we can approximate that
covariance structure with an $n$ x $K$ matrix where $K << n$. Note that we
discard any basis column that is zero at every observation, as it carries no
information for the model to split on, so $K$ counts only the retained columns.
Equally important, the resulting matrix can simply be concatenated with the
design matrix $X(s)$, treating the evaluated spatial basis as $K$ additional
covariates. This means no additional optimization is required to fit the model.
The default algorithm is sufficient.

```{r, results='hide'}
#| label: spatial-bases
#| code-fold: true
plt <- show_basis(basis_function) +
  geom_sf(
    data = aoi,
    fill = alpha(gvcols("orange"), 0.3),
    color = gvcols("orange"),
    linewidth = 1
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin()
  )

bb8 <- st_bbox(sites)

dx <- bb8[["xmax"]] - bb8[["xmin"]]
dy <- bb8[["ymax"]] - bb8[["ymin"]]

ggsave(
  filename = "figures/spatial-bases.svg",
  plot = plt,
  width = 4,
  height = 4 * (dy / dx)
)

remove(plt, bb8, dx, dy)
```

::: {#fig-spatial-bases style="width: 50%; margin: 0 auto;"}
![](/figures/spatial-bases.svg){width="100%"}

Range of support and knot (or center) of each Matérn32 radial basis function.
:::

### Collect data

Now we collect the explanatory and spatial basis features into a separate table.

```{r}
#| label: training-data
training_data <- sites |>
  st_drop_geometry() |>
  select(id, pa, period, streams, roads, lake) |>
  mutate(
    pa = factor(pa),
    period = factor(period)
  ) |>
  bind_cols(spatial_bases)

remove(spatial_bases)
```

## Analysis

To account for class-imbalance in the presence and background sites, we use a
down-sampled random forest as recommended Valavi *et al* (2021). This involves
limiting the training data for each tree in the forest to an equal number of
presence and background sites. The large number of trees, in this case N=1000,
ensures proper sampling of the background environment without biasing the
ensemble toward those background sites.

### Down-sampling utilities

We want to ensure that there are an equal number of PaleoIndian, Arhcaic, and
background sites in every tree of the random forest.

```{r}
#| label: down-sampling-funs
case_weights <- function(x) {
  classes <- paste(
    x[["period"]],
    x[["pa"]],
    sep = "-"
  )
  tbl <- table(classes)
  w <- tbl["PaleoIndian-1"] / tbl[classes]
  w <- unname(w)
  i <- which(classes == "PaleoIndian-1")
  w[i] <- 1
  w
}

sample_fraction <- function(x) {
  classes <- paste(
    x[["period"]],
    x[["pa"]],
    sep = "-"
  )
  k <- length(unique(classes))
  (k * sum(classes == "PaleoIndian-1")) / nrow(x)
}
```

This is roughly equivalent to

```{r}
#| label: down-sample-example
set.seed(SEED)

local({
  classes <- paste(
    training_data[["period"]],
    training_data[["pa"]],
    sep = "-"
  )

  class_sample <- sample(
    classes,
    size = sample_fraction(training_data) * nrow(training_data),
    replace = TRUE,
    prob = case_weights(training_data)
  )

  table(class_sample)
})
```

### Residual spatial autocorrelation

We want to test for spatial autocorrelation in the residuals, so we first fit
the full model and then check its residuals.

```{r}
#| label: random-forest
rf_base <- ranger(
  pa ~ lake + roads + streams + period,
  data = training_data,
  num.trees = N_TREES,
  probability = TRUE,
  case.weights = case_weights(training_data),
  sample.fraction = sample_fraction(training_data),
  num.threads = N_THREADS,
  seed = SEED
)
```

Check residuals

```{r}
#| label: morans-i-test
moran_test <- function(.geometry, .residuals) {
  suppressWarnings({
    neighbors <- st_dist_band(.geometry, lower = 0)

    spatial_weights <- st_kernel_weights(
      nb = include_self(neighbors),
      geometry = .geometry,
      kernel = "gaussian"
    )

    global_moran_perm(
      .residuals,
      nb = neighbors,
      wt = spatial_weights
    )
  })
}

observed <- with(training_data, as.numeric(levels(pa))[pa])

estimate <- predict(
  rf_base,
  data = training_data,
  type = "response",
  num.threads = N_THREADS,
  verbose = FALSE
)[["predictions"]][, "1"]

moran_test(
  st_geometry(sites),
  .residuals = observed - estimate
)
```

### Leave-one-out Cross Validation (LOOCV)

The AUC is a common metric used for evaluating the predictive power of SDMs, but
it summarizes a model over an entire test set, which tells us nothing about
individual sites. Because we want to interrogate model behavior on a per site
basis, we calculate instead the percentile rank of each left out PaleoIndian site
$s^*$ relative to all $i = 1, ..., M$ background points

$$
R = \frac{1}{M}\sum^{M}_{i=1} \mathbf{1}[\hat{p}(i) < \hat{p}(s^*)]
$$

with $\mathbf{1}[\cdot]$ being the indicator function, which is one when
$\hat{p}(i) < \hat{p}(s^*)$ and zero otherwise, and $\hat{p}$ being the estimate
returned by the model. This measure ranges over the unit interval $[0,1]$ and
reads as the proportion of the landscape that the focal site outscores, so
$R=0.5$ means the model cannot distinguish the site from a randomly chosen
location and $R=1$ means it ranks the site above every background point.

Estimating this for each PaleoIndian site requires looping through the data,
training a model on all PaleoIndian sites except a focal site (the leave-one-out
site), then estimating $\hat{p}$ for the focal site and for all background sites.
To simplify the for-loop, we define a helper function for fitting a model to
training data and evaluating it against test data.

::: callout-note
Because we are modeling presence-only data, $\hat{p}$ is more accurately
interpreted as a *relative likelihood* rather than a strict probability.
:::

```{r}
#| label: evaluate-fn
fit <- function(.formula, .train) {
  ranger(
    as.formula(.formula),
    data = .train,
    num.trees = N_TREES,
    probability = TRUE,
    case.weights = case_weights(.train),
    sample.fraction = sample_fraction(.train),
    num.threads = N_THREADS,
    seed = SEED
  )
}

estimate <- function(.model, .data) {
  yhat <- predict(
    .model,
    data = .data,
    type = "response",
    num.threads = N_THREADS,
    verbose = FALSE
  )
  yhat <- yhat[["predictions"]][, "1"]
  list(presence = yhat[1], absence = yhat[-1])
}

evaluate <- function(.estimate) {
  presence <- .estimate[["presence"]]
  absence <- .estimate[["absence"]]
  data.frame(
    rank = mean(absence < presence),
    difference = mean(presence - absence)
  )
}
```

Here we evaluate the LOO-CV metric for four models:

1.  Base: cost-distance
2.  Spatial: Base + spatial basis functions
3.  Comparative: Base + period
4.  Full: Base + spatial basis functions + period

Now we specify our models

```{r}
#| label: formulas
sbf_variables <- paste(
  grep("^sbf_", names(training_data), value = TRUE),
  collapse = " + "
)

formulas <- list(
  Base = "pa ~ streams + lake + roads",
  Spatial = paste0("pa ~ streams + lake + roads + ", sbf_variables),
  Comparative = "pa ~ streams + lake + roads + period",
  Full = paste0("pa ~ streams + lake + roads + period + ", sbf_variables)
)

remove(sbf_variables)
```

And evaluate them

```{r}
#| label: loocv
ids <- with(training_data, id[pa == 1 & period == "PaleoIndian"])
evaluations <- vector(mode = "list", length = length(ids))
names(evaluations) <- ids
estimates <- evaluations

for (id in ids) {
  test <- bind_rows(
    training_data |> filter(id == {{ id }}),
    training_data |> filter(pa == 0, period == "PaleoIndian")
  )

  train <- training_data |> filter(id != {{ id }})

  i <- with(train, which(period == "PaleoIndian"))

  models <- list(
    fit(formulas[["Base"]], train[i, ]),
    fit(formulas[["Spatial"]], train[i, ]),
    fit(formulas[["Comparative"]], train),
    fit(formulas[["Full"]], train)
  )

  estimates_ <- lapply(models, estimate, .data = test)
  names(estimates_) <- names(formulas)

  evals_ <- lapply(estimates_, evaluate)
  evals_ <- do.call(rbind, evals_)
  evals_[["model"]] <- names(formulas)
  evals_[["id"]] <- id

  estimates[[id]] <- estimates_
  evaluations[[id]] <- evals_[c("id", "model", "rank", "difference")]
}

evaluations <- do.call(rbind, evaluations)
rownames(evaluations) <- NULL

remove(i, id, ids, train, test, models, evals_)
```

## Results

To help interpret the results of the analysis, we provide all of the following:

1.  LOOCV distributions across sites and models
2.  Accumulated Local Effects (ALEs)
3.  Spatial predictions

These are evaluated for each of the four models detailed above.

### Leave-one-out Cross Validation

We evaluate the ability of each model to discriminate presence from background
locations as described above.

```{r}
#| label: figure-loocv
#| code-fold: true
dp <- evaluations |>
  mutate(
    id = sub("26LA", "", id),
    id = fct_reorder(factor(id), rank, max),
    model = fct_relevel(factor(model), "Full", "Comparative", "Spatial")
  )

segments <- dp |>
  group_by(id) |>
  summarize(x = min(rank), xend = max(rank)) |>
  ungroup()

plt1 <- ggplot() +
  geom_segment(
    data = segments,
    aes(x, id, xend = xend, yend = id),
    color = graph_gray,
    linewidth = 0.3
  ) +
  geom_point(
    data = dp,
    aes(rank, id, color = model),
    size = 2
  ) +
  scale_color_manual(
    values = gvcols("light-drab", "pink", "orange", "yellow"),
    guide = "none"
  ) +
  labs(x = "Percentile Rank [R]", title = "A. Site Distributions") +
  theme(axis.title.y = element_blank())

labels <- dp |>
  group_by(model) |>
  summarize(x = quantile(rank, probs = 0.25)) |>
  ungroup()

plt2 <- ggplot() +
  geom_boxplot(
    data = dp,
    aes(rank, model, fill = model),
    linewidth = 0.3
  ) +
  geom_point(
    data = dp,
    aes(rank, model),
    size = 2,
    shape = 21,
    color = "black",
    fill = "white",
    stroke = 0.2
  ) +
  geom_text(
    data = labels,
    aes(x, model, label = model, color = model),
    nudge_x = -0.02,
    nudge_y = 0.2,
    hjust = 1,
    vjust = 0,
    size = 12 / .pt
  ) +
  scale_fill_manual(
    values = gvcols("light-drab", "pink", "orange", "yellow")
  ) +
  scale_color_manual(
    values = gvcols("drab", "dark-pink", "dark-orange", "dark-yellow")
  ) +
  labs(x = "Percentile Rank [R]", title = "B. Model Distributions") +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none"
  )

plt <- plt1 + plt2 + plot_layout(axis_titles = "collect")

ggsave(
  filename = "figures/loocv-results.svg",
  plot = plt,
  width = 7,
  height = 4
)

remove(dp, segments, labels, plt, plt1, plt2)
```

::: {#fig-loocv-results style="width: 90%; margin: 0 auto;"}
![](/figures/loocv-results.svg){width="100%"}

Results of leave-one-out cross validation. (A) Distribution of percentile ranks
across models for each PaleoIndian site. (B) Distribution of precentile ranks 
across PaleoIndian sites for each model.
:::

**Spatial sorting bias.** Note that the above LOO-CV currently ignores potential
spatial sorting bias in the estimate, including the difference no matter the
distance between the target PaleoIndian site and the background location.
However, Tobler's Law implies that differences in probability at presence and
background locations should increase with distance, all else being equal. Here
we provide a visual check of this possibility, which also helps to interrogate
the behavior of the models.

```{r}
#| label: figure-loocv-distance
#| code-fold: true
D <- st_distance(
  sites |> filter(pa == 1, period == "PaleoIndian"),
  sites |> filter(pa == 0, period == "PaleoIndian")
)

D <- D / 1000
D <- units::drop_units(D)

ids <- with(sites, id[pa == 1 & period == "PaleoIndian"])

rownames(D) <- ids
colnames(D) <- with(sites, id[pa == 0 & period == "PaleoIndian"])

r <- seq(5, 65, by = 10)

for (i in seq_len(nrow(evaluations))) {
  id <- evaluations[["id"]][i]
  mod_ <- evaluations[["model"]][i]
  prank <- c()

  for (threshold in r) {
    j <- which(c(D[id, ]) <= threshold)
    j <- unname(j)

    p_1 <- estimates[[id]][[mod_]][["presence"]]
    p_0 <- estimates[[id]][[mod_]][["absence"]][j]

    prank <- c(prank, mean(p_0 < p_1))
  }

  evaluations[["range"]][i] <- list(prank)
}

M <- as.data.frame(do.call(rbind, evaluations[["range"]]))
colnames(M) <- r

M <- cbind(evaluations[c("id", "model")], M)

M <- M |>
  pivot_longer(c(-id, -model), names_to = "distance", values_to = "value") |>
  mutate(
    distance = as.numeric(distance),
    model = fct_relevel(factor(model), "Full", "Comparative", "Spatial")
  )

plt <- ggplot(M, aes(distance, value, group = model, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 0.8) +
  scale_color_manual(
    name = "Model",
    values = gvcols("light-drab", "pink", "orange", "yellow"),
    guide = guide_colorlegend(reverse = TRUE)
  ) +
  facet_wrap(vars(id)) +
  labs(x = "Distance [km]", y = "Percentile Rank [R]") +
  scale_x_continuous(breaks = seq(5, 65, by = 20)) +
  theme(
    legend.position = "top",
    legend.margin = margin(),
    legend.title = element_text(margin = margin(r = 6))
  )

ggsave(
  plot = plt,
  filename = "figures/loocv-distance.svg",
  width = 6.5,
  height = 7
)

remove(i, j, id, ids, r, p_1, p_0, D, M, plt)
```

::: {#fig-loocv-distance style="width: 80%; margin: 0 auto;"}
![](/figures/loocv-distance.svg){width="100%"}

Results of leave-one-out cross validation compared to background points at
different distances from each left out PaleoIndian site.
:::

### Accumulated Local Effects (ALEs)

```{r}
#| label: ales
ale <- function(variable, .model, .data) {
  x <- .data[[variable]]

  # bin the focal variable at its own quantiles
  breaks <- quantile(
    x,
    probs = seq(0, 1, length = N_GRID + 1),
    names = FALSE
  )
  breaks <- unique(breaks)
  bin <- .bincode(x, breaks, include.lowest = TRUE)

  # predict at each observation's bin edges, holding all else fixed
  lower <- .data
  upper <- .data
  lower[[variable]] <- breaks[bin]
  upper[[variable]] <- breaks[bin + 1]
  lower <- unlist(estimate(.model, lower), use.names = FALSE)
  upper <- unlist(estimate(.model, upper), use.names = FALSE)
  delta <- upper - lower

  # accumulate mean within-bin differences across bins
  bins <- factor(bin, levels = seq_along(breaks[-1]))
  mean_delta <- tapply(delta, bins, mean)
  mean_delta[is.na(mean_delta)] <- 0
  yhat <- cumsum(c(0, mean_delta))

  # center so the data-weighted average effect is zero
  yhat <- yhat - mean(yhat[bin + 1])

  data.frame(variable = variable, x = breaks, yhat = yhat)
}

i <- which(training_data[["period"]] == "PaleoIndian")

models <- list(
  Base = fit(formulas[["Base"]], training_data[i, ]),
  Comparative = fit(formulas[["Comparative"]], training_data),
  Spatial = fit(formulas[["Spatial"]], training_data[i, ]),
  Full = fit(formulas[["Full"]], training_data)
)

variables <- c("lake", "streams", "roads")

ales <- Map(
  ale,
  variable = rep(variables, each = 2),
  .model = models[c("Base", "Comparative")],
  MoreArgs = list(.data = training_data[i, ]),
  USE.NAMES = FALSE
)

model_labels <- rep(
  rep(c("Base", "Comparative"), times = 3),
  times = sapply(ales, nrow)
)

ales <- do.call(rbind, ales)

ales[["model"]] <- model_labels
```

```{r}
#| label: figure-ales
#| code-fold: true
ales[["model"]] <- fct_relevel(
  ales[["model"]],
  "Comparative"
)

plt <- ggplot(ales, aes(x, yhat, color = model)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(
    name = "Model",
    values = gvcols("pink", "yellow"),
    guide = guide_colorlegend(reverse = TRUE)
  ) +
  facet_wrap(
    vars(variable),
    ncol = 3,
    scales = "free_x",
    labeller = labeller(variable = \(.x) {
      paste("To", str_to_title(.x))
    })
  ) +
  labs(x = "Hours", y = "Accumulated Local Effect") +
  theme(
    legend.position = "bottom",
    legend.margin = margin(),
    legend.title = element_text(margin = margin(r = 6))
  )

ggsave(
  plot = plt,
  filename = "figures/ales.svg",
  width = 6.3,
  height = 2.9
)

remove(plt)
```

::: {#fig-ales style="width: 90%; margin: 0 auto;"}
![](/figures/ales.svg){width="100%"}

Accumulated local effects. Response is technically the centered relative 
likelihood of occurrence of a PaleoIndian site, not a strict probability.
:::

### Predictions

For spatial prediction, we need a raster for each covariate. We have that for
the explanatory variables, but the spatial basis functions require that we apply
Matérn32 to the distance from each grid cell to each knot, as we did with the
presence and background locations.

```{r}
#| label: spatial-covariates
# time period raster (the dummy variable)
r_period <- setValues(dem, "PaleoIndian")
names(r_period) <- "period"

# function to build a raster for each spatial basis function
sbf_to_raster <- function(x, r) {
  xy <- xyFromCell(r, cells(r))
  M <- eval_basis(x, xy)

  # build a raster for each knot (or centroid for spatial basis function)
  rr <- vector(mode = "list", length = ncol(M))
  for (i in seq_len(ncol(M))) {
    r[cells(r)] <- M[, i, drop = TRUE]
    rr[[i]] <- r
  }

  # stack the list of rasters into a single raster with multiple layers
  rast(rr)
}

r_spatial <- sbf_to_raster(basis_function, dem)
names(r_spatial) <- grep("^sbf_", names(training_data), value = TRUE)

# stack all covariate rasters and mask just to make sure we don't predict
# outside of the project area
r_covariates <- mask(c(r_period, costs, r_spatial), dem)

remove(r_period, sbf_to_raster, r_spatial)
```

Now we feed these covariate rasters into our prediction function.

```{r}
#| label: predictions
estimate <- function(.model, .data) {
  yhat <- predict(
    .model,
    data = .data,
    type = "response",
    num.threads = N_THREADS,
    verbose = FALSE
  )
  yhat[["predictions"]][, "1"]
}

r_predictions <- lapply(
  models,
  \(rf) {
    predict(r_covariates, rf, fun = estimate)
  }
)

r_predictions <- rast(r_predictions)
names(r_predictions) <- names(formulas)
```

```{r}
#| label: figure-predictions
#| code-fold: true
r_predictions <- mask(r_predictions, dem)

plt <- ggplot() +
  geom_spatraster_rgb(data = hill) +
  geom_spatraster(data = r_predictions, alpha = 0.5) +
  scale_fill_distiller(
    palette = "YlOrBr",
    direction = 1,
    na.value = "transparent"
  ) +
  facet_wrap(
    vars(lyr),
    nrow = 1
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    plot.margin = margin()
  )

ggsave(
  filename = "figures/predictions.svg",
  plot = plt,
  width = 7.5,
  height = 3.2
)
```

::: {#fig-predictions style="width: 100%; margin: 0 auto;"}
![](/figures/predictions.svg){width="100%"}

PaleoIndian response in geographic space.
:::

## Session Info

```{r}
#| label: session-info
# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh[["platform"]][["quarto"]] <- paste(
  quarto::quarto_version(),
  "@",
  quarto::quarto_path()
)

# print it out
pkg_sesh
```